---
title: "Tutorial Setup"
subtitle: "Instructions on how to work with the tutorials"
date: ""
toc: false
sidebar: false
format: html
---


Please make sure that you have followed the pre-course instructions before continuing with the instructions here.


## Working on Pelle (UPPMAX HPC)

We have booked 5 cores on Pelle per course participant. To run the tutorial in the interactive mode log on to Pelle and run interactive command:

```{r}
#| eval: false
ssh -Y <username>@pelle.uppmax.uu.se
```


While logged to Pelle, book the interactive session:
```{r}
#| eval: false
interactive -A uppmax2025-2-309 -p core -n 5 -t 08:00:00
```

This command gives you access to 5 cores for eight hours or until 17:00, whichever comes first. Please make sure you do it only ONCE per day, as by repeating this command you will use up resources reserved for others.

Check which node you were assigned
```{r}
#| eval: false
squeue -u <username>
```


And connect to your node with
```{r}
#| eval: false
ssh -Y <nodename>
```


If you lose connection to the server during a tutorial, please connect again to Pelle and do `ssh -Y <nodename>` again. You will reconnect to your set of cores.


## Working on your local machine

Instructions to follow.
