---
title: "Peak Annotation"
format: 
  html:
    toc: true
    toc-depth: 2
    fig-format: png
    mermaid:
      theme: default
date: "`r format(Sys.time(), '%B %d, %Y')`"
editor: 
  markdown: 
    wrap: 72
bibliography: references.bib
---


```{r echo=FALSE,include=FALSE,cache=FALSE}
library(knitr)

knitr::opts_chunk$set(echo = TRUE, 
                      collapse = TRUE, 
                      warning = FALSE, 
                      message = FALSE,
                      include=TRUE,
                      cache=FALSE,
                      out.width = "99%")

options(knitr.duplicate.label = "allow")

```

```{=html}
<!-- 
Author:         Agata Smialowska
About:          This qmd file is used to render the .html file. To modify or 
                update the .html file, please do so in the .qmd file here and 
                render it.
-->
```


<br />

## Learning outcomes


Using packages from Bioconductor:


* to profile ATAC-seq signal by genomic location and by proximity to TSS regions;

* to annotate peaks with the nearest feature;

* to perform functional annotation.


<br />
<br />


::: {.callout-note}

  
   We continue working with data from [@Tsao2022]. We will use the count table derived from non subset data (already prepared).

   We will also use results of [Differential Accessibility](../ATACda/index.html).

   Data processing recap: reads were mapped to reference genome `mm39`.

:::


<br />
<br />


## Introduction

In this tutorial we use an R / Bioconductor package `ChIPseeker` [@Yu2015], to inspect the ATAC signal profiles (can be used for ChIP-seq as well), annotate peaks and visualise annotations. We will also perform functional annotation of peaks using `clusterProfiler` [@Yu2012].

This tutorial is based on the [ChIPseeker package tutorial](https://bioconductor.org/packages/release/bioc/vignettes/ChIPseeker/inst/doc/ChIPseeker.html) so feel free to have this open alongside to read and experiment more.


<br />
<br />

## Data & Methods

We will build upon the main labs:

* ATAC-seq: all detected peaks (merged consensus peaks);

* ATAC-seq: differentially accessible peaks;

<br />
<br />

## Setting Up

You can continue working in the directory `atacseq/analysis/counts`. This directory contains merged peaks called earlier using genrich as well as count tables derived from summarising of non-subset data. Although strictly speaking we won’t need the count tables for this exercise, this will be our staring point, to prepare the object for the differential accessibility analysis. We will use file `AB_Batf_KO_invivo.genrich_joint.merged_peaks.featureCounts` and annotation libraries, which are preinstalled. 


::: {.callout-note}
  
   We take advantage of the module system on Pelle in this tutorial. The code was tested under `R 4.6.1` The lab was developed under the R version stated in session info.

:::

<br />
<br />

We navigate to our working directory for this tutorial. Assuming we are at the top of the `ataqseq/analysis`

```{bash}
#| eval: FALSE
#| echo: TRUE

cd counts
```

<br />
<br />


We access the R environment via:

```{bash}
#| eval: FALSE
#| echo: TRUE

module load R/4.6.1-gfbf-2026.1
module load R-bundle-CRAN/2026.07-foss-2026.1
module load R-bundle-Bioconductor/3.23-foss-2026.1-R-4.6.1
```

We activate R console upon typing `R` in the terminal.


To be able to load all necessary libraries we need to update the paths to custom library locations:


```{r}
#| eval: FALSE
#| echo: TRUE

.libPaths(c(.libPaths(),"/proj/epi2026/library/Rpackages/4.6.1"))
```

Alternatively you can work on this tutorial locally, using R libraries installed via `renv`.


We begin by loading necessary libraries:


```{r}
#| label: libraries
#| echo: TRUE

library(tidyverse)
library(dplyr)
library(kableExtra)

library(GenomicRanges)

library(ChIPseeker)
library(ggupset)
library(ggimage)
library(biomaRt)
library(txdbmaker)
library(GenomicFeatures)

library(org.Mm.eg.db)
library(fgsea)
library(ReactomePA)
```
<br />
<br />


<br />
<br />

## Preparing Annotations

We will use TSS and gene annotations from Ensembl. We can fetch them from `biomaRt`:


::: {.callout-note collapse="true"}

### Obtaining the annotation objects

```{r}
#| eval: false
#| echo: true

txdb_mm39 = txdbmaker::makeTxDbFromBiomart(biomart="ensembl",
                               dataset="mmusculus_gene_ensembl",
                               circ_seqs=NULL,
                               taxonomyId=NA)

ensembl = useEnsembl(biomart="genes", dataset="mmusculus_gene_ensembl")
all_genes = names(transcriptsBy(txdb_mm39, "gene"))

gene_annot_mm39=getBM(attributes = 
   c('ensembl_gene_id', 'ensembl_transcript_id','external_gene_name','entrezgene_id','description','gene_biotype','chromosome_name','start_position','end_position','strand','transcription_start_site'),
       filters = 'ensembl_gene_id',
       values = all_genes, 
        mart = ensembl)
```

:::

<br />
<br />

We have prepared these annotation files for you, so you can load them via:

<br />

```{r}
#| eval: false
#| echo: true

annotdir=file.path(getwd(),"annotation")

gene_annot_mm39=read.delim(file.path(annotdir,"mm39_gene_names.tab"), sep="\t", header=TRUE, quote = "")
```


```{r}
#| echo: false

annotdir=file.path("assets","annotation")

gene_annot_mm39=read.delim(file.path(annotdir,"mm39_gene_names.tab"), sep="\t", header=TRUE, quote = "")
```

<br />

and

<br />

```{r}
#| eval: FALSE
#| echo: true

pth_txdbens=file.path(annotdir,"Ensembl.txdb.GRCm39.Rdata")
txdb_mm39=loadDb(pth_txdbens)
```


<br />
<br />


You can inspect the objects:


```{r}
#| eval: FALSE
#| echo: true

txdb_mm39
## TxDb object:
## # Db type: TxDb
## # Supporting package: GenomicFeatures
## # Data source: BioMart
## # Organism: Mus musculus
## # Taxonomy ID: 10090
## # Resource URL: www.ensembl.org:443
## # BioMart database: ENSEMBL_MART_ENSEMBL
## # BioMart database version: Ensembl Genes 115
## # BioMart dataset: mmusculus_gene_ensembl
## # BioMart dataset description: Mouse genes (GRCm39)
## # BioMart dataset version: GRCm39
## # Full dataset: yes
## # miRBase build ID: NA
## # Nb of transcripts: 278396
## # Db created by: txdbmaker package from Bioconductor
## # Creation time: 2025-09-08 11:35:27 +0200 (Mon, 08 Sep 2025)
## # txdbmaker version at creation time: 1.0.1
## # RSQLite version at creation time: 2.4.3
## # DBSCHEMAVERSION: 1.2
```


```{r}
#| collapse: TRUE

head(gene_annot_mm39)
```

<br />
<br />




::: {.callout-note collapse="true"}

### Working with UCSC annotations
  
If you would rather annotate TSS using the gene models from UCSC you can use the Bioconductor package directly:

```
library(TxDb.Mmusculus.UCSC.mm39.knownGene)
txdb = TxDb.Hsapiens.UCSC.mm39.knownGene
```

Please note that UCSC and Ensembl use different contig naming schemes, so it is advisable to use the annotation matching the genome reference used for read mapping.

:::

<br />
<br />



## Data

We can now load data: 

```{r}
#| eval: FALSE
#| echo: true

count_table_fname="AB_Batf_KO_invivo.genrich_joint.merged_peaks.featureCounts"
cnt_table=read.table(count_table_fname, sep="\t", header=TRUE, blank.lines.skip=TRUE)
```

```{r}
#| eval: true
#| echo: false

count_table_fname="AB_Batf_KO_invivo.genrich_joint.merged_peaks.featureCounts"
cnt_table_pth=file.path("assets","data",count_table_fname)
cnt_table=read.table(cnt_table_pth, sep="\t", header=TRUE, blank.lines.skip=TRUE)
```

<br />
<br />

We will format the column names and subset the count table to only contain the peaks on assembled chromosomes.


```{r}
rownames(cnt_table)=cnt_table$Geneid
rownames(cnt_table)=c(gsub("AB_Batf_KO_invivo.genrich_joint.","",rownames(cnt_table)))
colnames(cnt_table)=c(colnames(cnt_table)[1:6],gsub(".filt.bam","",colnames(cnt_table)[7:10]))

colnames(cnt_table)[7:10]=c("B1_WT_Batf-floxed","B2_WT_Batf-floxed","A1_Batf_cKO","A2_Batf_cKO")

#remove peaks not on the assembled chromosomes
cnt_table_chr=cnt_table|>
  dplyr::filter(Chr%in%c(1:19) | Chr%in%c("X","Y"))

reads.peak=cnt_table_chr[,c(7:10)]

head(reads.peak)
```


* All peaks: n = `r nrow(cnt_table)`;

* Peaks on assembled chromosomes: n = `r nrow(cnt_table_chr)`. These peaks will be used for further analysis.

<br />
<br />


## Peak Annotation

We will be working on a `GRanges` object `peaks_gr` containing non-subset peaks (i.e. all assebled chromosome peaks).

We create the `GRanges` object:

```{r}
peaks_gr=GRanges(seqnames=cnt_table_chr$Chr, ranges=IRanges(cnt_table_chr$Start, cnt_table_chr$End), strand="*", mcols=data.frame(peakID=rownames(cnt_table_chr)))
```

and inspect it:

```{r}
peaks_gr
```

<br />
<br />



We are ready to annotate the peaks to their closest feature:


```{r}
#| eval: FALSE
#| echo: true
peakAnno=annotatePeak(peaks_gr, tssRegion=c(-3000, 3000),TxDb=txdb_mm39)
```

```{r}
#| echo: FALSE
#| eval: FALSE
saveRDS(peakAnno, 
  file = "assets/rds/Batf_WT_KO.merged_peaks.annot.obj.rds", 
  ascii = FALSE, 
  version = NULL, 
  compress = TRUE, 
  refhook = NULL)
```

```{r}
#| eval: TRUE
#| echo: FALSE
peakAnno=readRDS("assets/rds/Batf_WT_KO.merged_peaks.annot.obj.rds")
```

Summary of the regions annotated to peaks:


```{r}
peakAnno
```

Over 30% peaks localised to TSS, as expected in an ATAC-seq experiment.

<br />
<br />


To inspect peak annotations:

```{r}
peakAnno_df=as.data.frame(peakAnno)
```


```{r}
head(peakAnno_df)
```

<br />
<br />


We may want to include more gene related information:

```{r}
peakAnno_df=peakAnno_df|>
  left_join(gene_annot_mm39, by=c("transcriptId"="ensembl_transcript_id"))|>
  dplyr::rename(peakID=mcols.peakID)
```




<br />
<br />


::: {.callout-note collapse="true"}

### Annotated peaks can be saved to a file as a table or rds object

```{r}
#| eval: FALSE
#| echo: true

write.table(peakAnno_df, "Batf_WT_KO.merged_peaks.tsv", 
    append = FALSE, 
    quote = FALSE, 
    sep = "\t",
    row.names = FALSE,
    col.names = TRUE, 
    fileEncoding = "")
```

<br />
<br />

```{r}
#| eval: FALSE
#| echo: false

saveRDS(peakAnno_df, 
  file = "assets/rds/Batf_WT_KO.merged_peaks.annot.rds", 
  ascii = FALSE, 
  version = NULL, 
  compress = TRUE, 
  refhook = NULL)
```


:::

<br />
<br />

We can also assess the read density at annotated TSS regions. The plot of read density at the transcription start sites (TSS) for all peaks is presented on @fig-tss-profile.

```{r}
#| eval: FALSE
#| echo: true

promoter=getPromoters(TxDb=txdb_mm39, upstream=3000, downstream=3000)
```

```{r}
#| eval: FALSE
#| echo: false

promoter=getPromoters(TxDb=txdb_mm39, upstream=3000, downstream=3000)
saveRDS(promoter, 
  file = "assets/rds/prom.ensembl.rds", 
  ascii = FALSE, 
  version = NULL, 
  compress = TRUE, 
  refhook = NULL)
```



```{r}
#| echo: FALSE
#| eval: true

promoter=readRDS("assets/rds/prom.ensembl.rds")
```


```{r}
tagMatrix=getTagMatrix(peaks_gr, windows=promoter)

TSS_profile=plotAvgProf(tagMatrix, xlim=c(-3000, 3000), xlab="Genomic Region (5'->3')", ylab = "Read Count in Peaks")
```



::: {#fig-tss-profile}

```{r}
TSS_profile
```
Plot of signal density in peaks in relation to transcription start sites (TSS). Expected is an enrichment of signal proximal to TSS.

:::



<br />
<br />


We can also plot summary of the annotations:

```{r}
peakAnnoplot=upsetplot(peakAnno, vennpie=TRUE)
```


::: {#fig-annot-upset}

```{r}
peakAnnoplot
```
Summary of features annotated to peaks.

:::


::: {.callout-note collapse="true"}

### Saving the plots


```{r}
#| eval: FALSE
#| echo: true

pdf("TSSdist.pdf")
  TSS_profile
dev.off()

pdf("AnnotVis.pdf")
  peakAnnoplot
dev.off()

```

:::


<br />
<br />

<br />
<br />

## Functional Analysis

Having obtained annotations to nearest genes, we can perform functional enrichment analysis to identify predominant biological themes among these genes by incorporating knowledge provided by biological ontologies, e.g. GO (Gene Ontology, [@GO]) and Reactome [@reactome].

<br />

We can use several approaches and peak sets for this, e.g.:

* genes annotated to merged consensus peaks set in e.g. promoters or proximal to TSS to assess functional signature in open chromatin regions (OCRs) (overrepresentation analysis set vs. id universe, ORA);

* differentially accessible peaks to define functional signature in OCRs which change their occupancy status (ORA);

* genes annotated to merged consensus peaks set ranked by an effect size metric (Gene Set Enrichment Analysis, GSEA).

<br />
<br />

### Pathway Over-Representation Analysis

<br />

Overrepresentation analysis (ORA) is a method to test whether genes within given gene set are overepresented in the **subset** (open chromatin, DA peaks, etc.) **vs. the universe** (all genes). 


<br />

Let’s first annotate the consensus peaks, representing open chromatin regions (OCRs) with **Reactome** pathways. 

Reactome uses **entrez gene ID** space for gene - pathway annotation.

<br />
<br />


Reactome pathway enrichment of genes defined as the nearest feature to the peaks:

```{r}
entrez_ids=peakAnno_df$entrezgene_id
entrez_ids=entrez_ids[!is.na(unique(entrez_ids))]

pathway.reac=ReactomePA::enrichPathway(entrez_ids, organism = "mouse")
```

<br />
<br />

```{r}
#previewing enriched Reactome pathways
colnames(as.data.frame(pathway.reac))

#we skip the preview of some columns which contain long strings of gene IDs
pathway.reac[1:10,c(1:7,9)]
```

<br />
<br />

We can see familar terms which can be connected to sample biology: 

* Cytokine Signaling in Immune system

* Class I MHC mediated antigen processing & presentation


<br />
<br />


### Gene Set Enrichment Analysis

<br />

Gene set enrichment analysis (GSEA) is a test whether gene set member genes are present towards to top of the **ranked list** of genes [@gsea]. The rank usually corresponds to the effect size and statistical significance. Ties in the ranked list are discouraged, as they may lead to unexpected results.

<br />

We can check for GO terms enriched amongst DA OCRs by using GSEA.


We need to read in the results of the DA analysis:

<br />

```{r}
#| eval: TRUE
#| echo: false

DA_res=readRDS("assets/rds/Batf_WT_KO.merged_peaks.DA_TMM.annot.rds")
```


```{r}
#| eval: false
#| echo: true

DA_res=readRDS("../../results/DA/objects/Batf_WT_KO.merged_peaks.DA_TMM.annot.rds")
```

<br />

The metric to use in GSEA ideally should combine effect size and statistical significance. 

Common practice for constructing the metric based on `edgeR` results is:

<br />

```{r}
DA_res=DA_res|>
  dplyr::mutate(gsea_metric=PValue*logFC)
```

<br />

GSEA statistical framework recommends that the ranked gene list is duplicate free; yet often a gene may have more than one peak annotated. If we leave the list as is (i.e. with duplicated gene ids and different value for ranking metric for each entry), we have no control over which peak / metric value gets selected for GSEA, which may lead to unexpected results. Applying some selection strategy for genes with multiple peaks helps with interpretability of the GSEA results.

<br />
<br />

In this tutorial, we select:

* genes with ATAC peaks up to 3 kb of their TSS (annotated as `Promoter`);

* one peak per gene; we select the peak with largest absolute fold change;


This will require some data wrangling:

```{r}
DA_res.prom.top=DA_res|>
  dplyr::mutate(gene_name=coalesce(external_gene_name,geneId))|>
  dplyr::filter(annotation%in%c("Promoter"))|>
  dplyr::group_by(gene_name)|>
  dplyr::mutate(gene_max_abs_lfc=max(abs(logFC)))|>
  dplyr::ungroup()|>
  dplyr::rowwise()|>
  dplyr::filter(abs(logFC)==gene_max_abs_lfc)|>
  dplyr::distinct(gene_name, .keep_all=TRUE)|>
  as.data.frame()

## we also need to add entrez gene id and clean up entries without it
DA_res.prom.top=DA_res.prom.top|>
    left_join(gene_annot_mm39[,c(2,4)], by=c("transcriptId"="ensembl_transcript_id"))|>
    dplyr::distinct(entrezgene_id, .keep_all=TRUE)|>
    dplyr::filter_at(vars(entrezgene_id), all_vars(!is.na(.)))

```



<br />
<br />

We can now create a *named list* (names are entrez gene id) and sort it:

```{r}
#| eval: false
#| echo: true

gene_list=DA_res.prom.top$gsea_metric #using pval*logFC
names(gene_list)=DA_res.prom.top$entrezgene_id
gene_list=gene_list[order(gene_list,decreasing=TRUE)]

head(gene_list)
```

<br />

We prepare the GO terms and run GSEA:

<br />

```{r}
#| eval: false
#| echo: true

all_go.bp=as.data.frame(org.Mm.egGO)|>
  dplyr::filter(Ontology=="BP")

go_terms=as.list(org.Mm.egGO2ALLEGS)
go_terms=go_terms[names(go_terms)%in%all_go.bp$go_id]

gsea_go=fgsea(pathways = go_terms, 
                  stats    = gene_list,
                  minSize  = 15,
                  maxSize  = 500,
                  scoreType    = "std")
```

<br />
<br />

The resulting object contains basic statistics of the GSEA test.

```{r}
#| eval: false
#| echo: true

 head(gsea_go)
     pathway      pval      padj    log2err         ES        NES size
1 GO:0000018 0.9928571 1.0000000 0.05400883  0.1824006  0.6768697  120
2 GO:0000027 0.9490446 1.0000000 0.05080541  0.2415674  0.6675675   24
3 GO:0000028 0.1902072 0.8391112 0.13649044 -0.4844570 -1.2474683   19
4 GO:0000038 0.2611111 0.8800840 0.11284336 -0.4165674 -1.1610107   27
5 GO:0000041 0.3404255 0.9065779 0.09374654 -0.2997705 -1.0499786   83
6 GO:0000045 0.3294118 0.8975844 0.11331291  0.2753809  1.0549553  137
   leadingEdge
1 19264, 7....
2 72831, 2....
3 67097, 9....
4 14081, 1....
5 226025, ....
6 240327, ....

```

<br />

We can format these results:

<br />

```{r}
#| eval: false
#| echo: true

gsea_go=as.data.frame(gsea_go)

library(GO.db)
go_terms=select(GO.db, columns=c("GOID","TERM"), keys=gsea_go$pathway, keytype="GOID")

gsea_go=gsea_go|>
    dplyr::filter(pval<0.01)|>
    dplyr::select(pathway,NES,pval,padj,size)|>
    left_join(go_terms, by=c("pathway"="GOID"))

> head(gsea_go)
     pathway       NES        pval      padj size                         TERM
1 GO:0002064 -1.539558 0.001013134 0.5754347  181  epithelial cell development
2 GO:0003014 -1.542754 0.004758923 0.5757116   85         renal system process
3 GO:0006814 -1.434938 0.006011068 0.5757116  171         sodium ion transport
4 GO:0006836 -1.473969 0.002957773 0.5754347  184   neurotransmitter transport
5 GO:0007431 -1.702995 0.003805549 0.5757116   31   salivary gland development
6 GO:0007435 -1.738422 0.002137247 0.5754347   27 salivary gland morphogenesis
```




Upon object inspection we can see terms related to immune cell biology, amongst other GO terms. 

* antigen processing and presentation (GO:0019882)

* antigen processing and presentation of peptide antigen (GO:0048002)

<br />
<br />

<br />
<br />

## Session Info

<br />
<br />


::: {.callout-note collapse="true"}


## Session Info


```{r}
#| label: session-info

sessionInfo()
```

:::

<br />
<br />

## References


