---
title: "Integrating scATAC-seq and scRNA-seq data"
format: 
  html:
    toc: true
    fig-format: png
date: "`r format(Sys.time(), '%B %d, %Y')`"
editor: 
  markdown: 
    wrap: 72
#bibliography: references.bib
---


```{r echo=FALSE,include=FALSE,cache=FALSE}
library(knitr)

knitr::opts_chunk$set(echo = TRUE, 
                      collapse = TRUE, 
                      warning = FALSE, 
                      message = FALSE,
                      include=TRUE,
                      cache=FALSE,
                      out.width = "99%")

options(knitr.duplicate.label = "allow")

```



## Learning outcomes


* Combine single cell ATAC-seq data with single cell RNA-seq data to better infer cell types.

<br />

## Setup

::: {.callout-note collapse="false"}

### Note on this workflow

  This tutorial is a continuation of [Single cell ATAC-seq](../scATAC/index.html). To run this on Pelle, setup and initialize R as you did in that exercise.

:::

::: {.callout-note}

Alternatively you can work on this tutorial locally if you have all the R packages loaded below installed on your computer.

:::


We begin by loading necessary libraries:

```{r}
#| label: libraries
#| eval: FALSE
#| echo: true

library(ggplot2)

library(Signac)
library(Seurat)

set.seed(1234)

```

<br />
<br />

## Load the data generated in the scATAC-seq exercise


All data for this tutorial is available on Pelle in the directory `/proj/epi2026/single_cell/scATAC_scRNA/`. If you work on the tutorial locally, you need to download the data and modify the path below according to your directory structure.

```{r}
#| eval: FALSE
#| echo: true

data.dir="/proj/epi2026/single_cell/scATAC_scRNA/"

```

The code here use a pre-calculated version. If you already have your pbmc object loaded, coming from the previous tutorial, you should skip running this code chunk.

::: {.callout-important collapse="false"}

  We assume that pbmc.Rda file you load was generated in the same file structure as you are running on now. That is the file structure you used in[scATAC](../scATAC/index.html) when you saved the file. Otherwise things will not work since Seurat Objects with fragments from ATAC-seq are not self contained, they contain paths to files with fragments and their indexes.

:::

```{r}
#| eval: FALSE
#| echo: true

load(file=paste0(data.dir, "data/pbmc.Rda"))
```

<br />
<br />

## Integrate with scRNA-seq data

In this section, we will compare the scATAC-seq data to scRNA-seq data from a similar starting material (although not from exactly the same cells). Since it is easier to figure out which cell types clusters correspond to in scRNA-seq data, we will then use the scRNA-seq data to annotate the clusters in the scATAC-seq data. (Note that if you have multi-omics data, with RNA-seq and ATAC-seq from the same cells, you don't need to take the extra detour to gene activity scores, but can instead use your RNA-seq data directly to annotate cell types.)

<br />

### Gene activity matrix

Gene activity scores capture how much open chromatin there is in the promoter regions of each gene (by default 2000 bp upstream). The assumption here is that open chromatin is a proxy for gene expression. Gene activity scores are represented as a matrix, with one row per gene and one column per cell. This makes the gene activity scores directly compatible with single cell RNA-seq data.

Calculating the gene activity scores can take around 10 minutes for 2000 cells and all genes.

```{r}
#| eval: FALSE
#| echo: true

DefaultAssay(pbmc) <- 'peaks'

gene.activities <- GeneActivity(pbmc)

# Add the gene activity matrix to the Seurat object as a new assay and normalize it
pbmc[['RNA']] <- CreateAssayObject(counts = gene.activities)
pbmc <- NormalizeData(
    object = pbmc,
    assay = 'RNA',
    normalization.method = 'LogNormalize',
    scale.factor = median(pbmc$nCount_RNA)
)
GetAssayData(pbmc$RNA)[1:10,1:3]
```

We can plot the gene activity scores for a set of genes across the clusters to get a first indication how it has worked.

```{r}
#| eval: FALSE
#| echo: true

DefaultAssay(pbmc) <- 'RNA'
FeaturePlot(
    object = pbmc,
    features = c('MS4A1', 'CD3D', 'LEF1', 'NKG7', 'TREM1', 'LYZ'),
    pt.size = 0.1,
    max.cutoff = 'q95',
    ncol = 3
)
```

<br />
<br />

### Label transfer

Having computed the gene activity scores, we are now ready to combine the ATAC-seq data with the RNA-seq data.

We start by finding anchors, *i.e.* pairs of cells, one from ATAC-seq and one from RNA-seq. This is done by first projecting ATAC-seq and RNA-seq data into the same space, and then find pairs of cells one from ATAC-seq and the other from RNA-seq that are mutual nearest neighbors (MNNs). These are further filtered and the reliable pairs are then used as anchors.

These anchors can then be used to project the ATAC-seq data onto the RNA-seq data and find the cell type annotation of the nearby cells. This way annotations from the RNA-seq data can be transferred to the ATAC-seq data. This is sometimes referred to as **label transfer**.

The corresponding RNA-seq data of human peripheral blood mononuclear cells (PBMCs) provided by 10x Genomics covers 19089 genes in 9432 cells.

```{r}
#| eval: FALSE
#| echo: true

# Load the pre-processed scRNA-seq data for PBMCs
pbmc_rna_v3 <- readRDS(paste0(data.dir,"data/pbmc_10k_v3.rds"))
pbmc_rna <- UpdateSeuratObject(pbmc_rna_v3)

# Find anchors
transfer.anchors <- FindTransferAnchors(
    reference = pbmc_rna,
    query = pbmc,
    reduction = 'cca'
)

predicted.labels <- TransferData(
    anchorset = transfer.anchors,
    refdata = pbmc_rna$celltype,
    weight.reduction = pbmc[['lsi']],
    dims = 2:30
)

pbmc <- AddMetaData(object = pbmc, metadata = predicted.labels)
```

```{r}
#| eval: FALSE
#| echo: true

plot1 <- DimPlot(
    object = pbmc_rna,
    group.by = 'celltype',
    label = TRUE,
    label.size = 3,
    repel = TRUE) + NoLegend() + ggtitle('scRNA-seq')

plot2 <- DimPlot(
    object = pbmc,
    group.by = 'predicted.id',
    label = TRUE,
    label.size = 3,
    repel = TRUE) + NoLegend() + ggtitle('scATAC-seq')

plot3 <- DimPlot(
    object = pbmc,
    label = TRUE,
    label.size = 3,
    repel = TRUE) + NoLegend() + ggtitle('scATAC-seq')

plot1 | plot2 | plot3

pbmc <- RenameIdents(
    object = pbmc,
    '0' = 'CD14 Mono',
    '1' = 'CD4/8 Naive',
    '2' = 'CD4 Memory (DN T)',
    '3' = 'CD14 Mono',
    '4' = 'CD8 Effector',
    '5' = 'pre-B/pro-B',
    '6' = 'NK dim/bright',
    '7' = 'Double neg. T',
    '8' = 'CD16 Mono',
    '9' = 'DC',
    '10' = 'pDC'
)
```

*How well do you think this worked? Can you see if some cells types are missing or merged?*

<br />

### Differentially accessible regions, again

Now, we will once again look at differentially accessible regions, this time between *CD4/8 Naive* cells and *CD14 Mono* cells (since we now can say what cell types these clusters represent). We can plot the signal in this region in a couple of different ways. 


```{r}
#| eval: FALSE
#| echo: true

# Change back to working with peaks instead of gene activities
DefaultAssay(pbmc) <- 'peaks'

# Find differentially accessible regions
da_peaks <- FindMarkers(
    object = pbmc,
    ident.1 = "CD4/8 Naive",
    ident.2 = "CD14 Mono",
    min.pct = 0.2,
    test.use = 'LR',
    latent.vars = 'peak_region_fragments'
)
head(da_peaks)
```

```{r}
#| eval: FALSE
#| echo: true

# Plot the signal for the most differentially accessible regions.
plot1 <- VlnPlot(
    object = pbmc,
    features = rownames(da_peaks)[1],
    pt.size = 0.1,
    idents = c("CD4/8 Naive","CD14 Mono")
)

plot2 <- FeaturePlot(
    object = pbmc,
    features = rownames(da_peaks)[1],
    pt.size = 0.1
)

plot1 | plot2
```

*Do you understand what these plots mean?*

<br />

### Annotating regions

Genomic coordinates are difficult to interpret on their own, and often it is interesting to know which genes are near or at the genomic regions. Below, for each region, we find the nearest gene and the distance to it. This makes it easy to check for regions near genes of interest, and to analyze genes further, *e.g.* through a Gene Ontology analysis.

```{r}
#| eval: FALSE
#| echo: true

open_cd4naive <- rownames(da_peaks[da_peaks$avg_log2FC > 0.5, ])
open_cd14mono <- rownames(da_peaks[da_peaks$avg_log2FC < -0.5, ])

closest_genes_cd4naive <- ClosestFeature(pbmc, regions = open_cd4naive)
closest_genes_cd14mono <- ClosestFeature(pbmc, regions = open_cd14mono)
head(closest_genes_cd4naive)
head(closest_genes_cd14mono)
```


<br />
<br />

## Plotting genomic regions

It is often informative to plot the ATAC-seq signal for a particular genomic region, and group this by cluster, cell type or some other meta data. We can either plot the aggregate signal for a group with `coveragePlot` or the signal for individual cells, with `tilePlot`. More info about plotting genomic region can be found [here](https://satijalab.org/signac/articles/visualization.html).

For plotting purposes, it’s nice to have related cell types grouped together. We can automatically sort the plotting order according to similarities across the annotated cell types by running the SortIdents() function:

```{r}
#| eval: FALSE
#| echo: true
#| 
pbmc <- SortIdents(pbmc)
```

```{r}
#| eval: FALSE
#| echo: true

CoveragePlot(
    object = pbmc,
    region = rownames(da_peaks)[1],
    extend.upstream = 40000,
    extend.downstream = 40000,
    peaks=TRUE
)
```

```{r}
#| eval: FALSE
#| echo: true

TilePlot(
    object = pbmc,
    region = rownames(da_peaks)[1],
    idents = c('CD14 Mono', 'CD4/8 Naive'),
    extend.upstream = 40000,
    extend.downstream = 40000
)
```


You can also highlight the differentially accessible regions overlapping a gene of interest.

```{r}
#| eval: FALSE
#| echo: true

# find DA peaks overlapping gene of interest
regions_highlight <- subsetByOverlaps(StringToGRanges(open_cd4naive), LookupGeneCoords(pbmc, "HOOK1"))

CoveragePlot(
  object = pbmc,
  region = "HOOK1",
  region.highlight = regions_highlight,
  extend.upstream = 1000,
  extend.downstream = 1000
)
```


<br />

One can also create an interactive version of these plots using the `CoverageBrowser()` function. One can use `CoverageBrowser()` to browser the genome and adjust plotting parameters interactively. The “Save plot” button in `CoverageBrowser()` will add the current plot to a list of `ggplot` objects that is returned when the browser session is ended by pressing the “Done” button, allowing interesting views to be saved during an interactive session.
