---
title: "Introduction to ChIP-seq Analyses"
author: "Dania Machlab"
format: 
  html:
    toc: true
    fig-format: png
date: "`r format(Sys.time(), '%B %d, %Y')`"
editor: 
  markdown: 
    wrap: 72
---


## Overview




```{r}
.libPaths(c("/Users/daniamachlab/Library/R/arm64/4.6-Bioc-3.23/library/", .libPaths()))

suppressPackageStartupMessages({
  library(QuasR)
  library(limma)
  library(BSgenome.Mmusculus.UCSC.mm39)
  library(GenomicRanges)
  library(ggplot2)
  library(tidyr)
  #library(GenomeInfoDb)
})

```


## Useful Bioconductor data structures

Throughout this lab we will be making use of a few data objects and packages from Bioconductor that are convenient when working with genomic data. They are described here and additional links are provided should you want to explore more or get a better understanding.

### GRanges

### BSgenome objects

### DNAstringSet

## Mapping


Download bam files from Pelle.

```{bash}
#| eval: false

rsync [username]@pelle.uppmax.uu.se:[path_to_file]

## Download the bam files
mkdir data
rsync danmac@pelle.uppmax.uu.se:'/proj/uppmax2026-1-197/ChIP/dataMapped/*_chr19.ucsc.fixed.bam*' data/ 
  
## Download the reference genome used
#mkdir reference
#rsync danmac@pelle.uppmax.uu.se:/proj/uppmax2026-1-197/ChIP/reference/Mus_musculus.GRCm39.dna.primary_assembly.fa reference/ 
```


## Quality Control (QC)

Let's begin by doing some QC on the mapped reads. We will be using the [QuasR](https://bioconductor.org/packages/QuasR) package to do this. 

```{r}
## Create the sample sheet needed by QuasR
## ... See ?qAlign for more details
bamFiles <- file.path(getwd(), 
                      list.files("data", pattern = ".bam$", full.names = TRUE))
meta <- data.frame(FileName = bamFiles, 
                   SampleName = limma::strsplit2(basename(bamFiles), "_chr19")[, 1])
write.table(meta, file = "data/sampleSheet.txt", quote = FALSE, sep = "\t", 
            row.names = FALSE, append = FALSE)

## Create proj object on aligned bam files
## ... you can set the number of cores here according to what works on you machine
cl <- makeCluster(4)
proj <- qAlign(sampleFile = "data/sampleSheet.txt", 
               genome = "BSgenome.Mmusculus.UCSC.mm39", 
               paired = "no", 
               clObj = cl, 
               checkOnly = TRUE)

## QC plots
dir.create("figures")
qQCReport(input = proj,  pdfFilename = "figures/mappingQC.pdf", clObj=cl)
```

Examine the generated `pdf` file and the plots it contains. We like to see high mapping rates typically close to 95% and no big differences in duplicate levels between the samples. Note that here you will see 100% since the bam files have already been filtered for only reads mapping to chr19.

### GC Bias

An important plot we often look at in ChIP-seq data is the GC-bias trends the samples have. Do they have very different ones and what are the consequences of that?

```{r}
## BSgenome
BSgenome <- BSgenome.Mmusculus.UCSC.mm39

## Tile the genome (on chr19 in our case) into 100 bp regions
tiles <- tileGenome(seqlengths = seqlengths(BSgenome)["chr19"], 
                    tilewidth = 500, 
                    cut.last.tile.in.chrom = TRUE)
tiles
```

Next, we will make use of the BSgenome object to extract the reference genome sequences and calculate the GC content of each tile.

```{r}

## Extract the raw reference sequences
tileSeq <- getSeq(BSgenome, tiles)

## Calculate G+C and CpG obs/expected
fMono <- oligonucleotideFrequency(tileSeq, width = 1L, as.prob = TRUE)
fDi <- oligonucleotideFrequency(tileSeq, width = 2L, as.prob = TRUE)
fracGC <- fMono[, "G"] + fMono[, "C"]
oeCpG <- (fDi[, "CG"] + 0.01) / (fMono[, "G"] * fMono[, "C"] + 0.01)

## Add the information to the tiles object
tiles$fracGC <- fracGC
tiles$oeCpG <- oeCpG

## Mark tiles with overlap with CpG islands
# url <- "https://hgdownload.cse.ucsc.edu/goldenPath/mm39/database/cpgIslandExt.txt.gz"
# url <- "https://hgdownload.cse.ucsc.edu/goldenPath/mm39/database/cpgIslandExt.txt.gz"
# 
# con <- pipe(paste("curl -L", shQuote(url)), "rb")
# cpg <- read.delim(gzcon(con), header = FALSE)
# close(con)
# 
# cpgIslands <- read.delim("https://hgdownload.cse.ucsc.edu/goldenPath/mm39/database/cpgIslandExt.txt.gz", 
#                          header = FALSE)

```


::: {.callout-tip}
## Exercise
Examine the different objects above and get more familiar with what they show. You can use the help pages of the used functions with `?functionName` ro inspect that the functions does, expects and gives back.
:::

We can now proceed to counting the reads on the tiles we have produced. This will take a few seconds.
```{r}
## Quantify reads
cntTiles <- qCount(proj = proj, query = tiles, mapqMin = 42, clObj = cl)
head(cntTiles)
```

Let's plot the log2(counts) vs the GC content of the tiles.

```{r}

## Prepare to plot
df <- as.data.frame(cbind(cntTiles, mcols(tiles))) %>%
  as_tibble() %>%
  pivot_longer(cols = starts_with("ChIPseq_"), names_to = "sample", values_to = "rawCount")

## Plot
ggplot(df, aes(x = fracGC, y = log2(rawCount+1))) +
  geom_point() + 
  facet_wrap(~sample)
  
  
# df <- as.data.frame(cbind(cnt_tiles, mcols(tiles)))
# set.seed(123)
# s <- sample(x = 1:nrow(df), size = 25000, replace = FALSE)
# df <- df[s, ]
# df <- df %>%
#   as_tibble() %>%
#   pivot_longer(cols = 2:10, names_to = "Sample", values_to = "raw_cnt")
# 
# ggplot(df) + 
#   geom_point(aes(x = percGC, y = log2(raw_cnt + 8), color = CGI.hasOverlap), alpha = 0.2) + 
#   geom_smooth(data = df[!df$CGI.hasOverlap, ], mapping = aes(x = percGC, y = log2(raw_cnt + 8)), span = 0.7) + 
#   facet_wrap(~ Sample) + 
#   theme_minimal()  
#   

```



mapping QC, GC bias

## Estimating fragment size

rsync danmac@pelle.uppmax.uu.se:'/proj/uppmax2026-1-197/ChIP/test/ChIPseq_ESC_CTCF_input_BPTFko_rep1_chr19.ucsc.bam*' data/ 

/home/danmac/Teaching/uppmax2026-1-197/ChIP/test/ChIPseq_ESC_CTCF_input_BPTFko_rep1_chr19.ucsc.bam


illustrate where TF sits and where true signal is, autocorrelation

## Peak calling
 
TF vs histone mark
 






